The Dynamics of Anthropomorphism in Human-Robot Interaction

### 2014-04-01 paper.tex
small changes in section 2

\section{Anthropomorphism}
\label{sec:anthropomorphism}

Anthropomorphism is an interesting phenomenon because it tells us something about 
the relationship between the human who anthropomorphizes, and the non-human agent 
which is anthropomorphized. Though commonly observed, it still appears controversial, at 
the same time, that humans tend to \textit{``see human in the non-
human''} \citep{epley_seeing_2007}. 
To shed light on this phenomenon, first of all, we would like to propose a clarification of terms.
In short, we propose to differentiate between \textbf{anthropomorphic form} to refer to an imitation of human-like form in the \textit{design} of artificial agents, and  \textbf{anthropomorphism} to refer to the social phenomenon of 
people's tendency to \textit{perceive} human-likeness in non human-like agents.
\citep{fink_anthropomorphism_2012} In the following, we might use the terms \texit{form} and \textit{design} interchangeably.

As anthropomorphism arises in an interaction between a human and a non-human agent
\citep{persson_anthropomorphism_2000}, one can also
take two main perspectives on explaining anthropomorphism \citep{lee_human_2005}: an artifact-centered perspective and a human-centered perspective. 
We now first present an artifact-centered perspective, and outline how 
robotics uses \textit{anthropomorphic design} to encourage 
people to interact in a social, human-like way with robots.
Then, we present a human-centered 
perspective and shed light on \textit{anthropomorphism as a social phenomenon}. We present 
theories from psychology and cognitive science to explain people's tendency to 
perceive human-like characteristics in non-human agents.
It has to be kept in mind that both perspectives are inseparable; they apply at the 
same time, and we would like to present them here just as two different viewpoints 
on the same phenomenon.
In the end of this section, we synthesize 
the presented backgrounds into an extended understanding of anthropomorphism 
as a multi-factor phenomenon.

@@

On one hand, developers and designers have discovered that human-like robots may better 
match the affordances that this to humans adapted world brings along. 
Apart from this, it is suggested that an increased human-likeness of a robot (up to a 
certain threshold) will generally lead to 
more perceived \textit{affinity} from the human side \citep{mori_uncanny_1970} (see 
Figure \ref{fig:uncanny_valley}).
This further increases anthropomorphism, and as such
appears to be beneficial for human-robot relations, especially with socially 
interactive robots \citep{fong_survey_2003}. Consequently, the domain of (social) 
robotics tries to exploit this by using
anthropomorphic forms in robots.
Humans seem to readily accept and prefer the 
anthropomorphic form of a robot (up to a certain point), when compared to purely mechanical 
robots. A possible explanation for this is that perceiving a system as
human-like (or also pet-like) compared to machine-like changes the 
way how users try to understand and make sense of this system. The reason for this is 
that humans tend to project everyday 
\hl{expectations} of human and animalistic life onto artificial agents that use 
anthropomorphic (or zoomorphic) forms \citep{schmitz_concepts_2011}. 
Both cognitive science 
and social psychology can provide explanations for this process of projection. The basic idea 
is that \textit{"the more similar in 
appearance, the more people are likely to use themselves as a source of induction 
and anthropomorphize these nonhuman agents"} \citep{epley_seeing_2007}. Studies have further 
shown that perceived similarity between the user and the robot influences the intensity of 
the participants' attributions towards the robot. Thus, the anthropomorphic design of a robot 
impacts the user's perception of the robot, and how far the user is likely to attribute and 
experience qualities such as intelligence, fun, or intentions in the robot. 
These experiences mark the transition between an immediate \textit{in-the-moment} and more thoughtful \textit{reflective} \citep{takayama_perspectives_2012} perception of, and interaction with the robot. As such, a more human-centered perspective needs to be applied, and we will come back to this in Section \ref{sec:human-centered}.

@@

Both the second factor \textit{effectance} and the third factor \textit{sociality} are \textbf{motivational determinants of anthropomorphism}. \textit{Effectance} is understood a person's motivation to interact effectively in one's environment. That means, a person is motivated to be able to understand, predict, and reduce uncertainty \footnote{\hl{(this feeds again our hypothesis that as soon as a robot is not uncertain anymore, there is no need to keep on anthropomorphizing it)}}\fxfatal{important! move/copy that to the 'model' section or discussion} about her environment and the agents that inhabit it. According to \cite{epley_seeing_2007}, \textit{effectance motivation} can lead to anthropomorphism, since anthropomorphism serves as one way to reduce uncertainty and to increase comprehension of events in one's environment. Again, also according this factor it can be argued that children are more likely to anthropomorphize than adults. As children are in their early stages of life, they are likely to feel more uncertain within their environment, and consequently tend to anthropomorphize more.

@@

To sum it up, while anthropomorphism is a commonly discussed and
studied trait of human-robot interaction, its the understanding 
does not seem to take into account the wide range of phenomena that it
encompasses. To date, the understanding of anthropomorphism in robotics considers  two main factors: the
characteristics of \textbf{1) the robot's design} (degree of human-likeness) and
the psychological determinants in \textbf{2) the human user} (see
\cite{epley_seeing_2007}). A lot of research has shown that both these factors can
promote or hinder anthropomorphism.

However, based on what we have just presented, we would like to make a note, that there are more than these two factors that impact on anthropomorphism.

Based on the understanding that anthropomorphism represents a specific type of \textit{experience} that arises in the
\textit{interaction} between a set of user expectations and external reality
\cite{persson_anthropomorphism_2000}, 
%(see Figure~\ref{fig:anthropomorphism_and_interaction})
we propose that
anthropomorphism is a multi-layered phenomenon which arises in different levels, and is likely to evolve with continued interaction.
%we suggest to understand anthropomorphism as a \emph{dynamic},
%\emph{context-dependent} process. To this end, we apply Persson \textit{et
%al.}'s six levels on anthropomorphism to HRI and introduce three
%\emph{interaction phases}. These interaction phases of anthropomorphism reflect
%also the results of a long-term study that we conducted in a real human
%environment and provide an account of the so-called \textit{novelty effect}.

To this end, we make the following two suggestions:

@@

Second, we would like to take up the proposition already made by other researchers (\eg \cite{persson_anthropomorphism_2000,ruijten_introducing_2014}) to not viewing anthropomorphism as being either there or not there, but to discriminate different levels of
anthropomorphism (low, and high anthropomorphism, for instance).
This seems to make sense, since each of the four proposed factors has its own characteristics and involves specific
types of user expectations, for instance. Such a differentiation in the understanding of anthropomorphism could further allow to draw inferences
on the human-robot relationship based on the observed level of anthropomorphism.

@@

@@

Why do we think that anthropomorphic projections are likely to change over time
and with growing experience a person has with a robot? The underlying process in
anthropomorphism is understood as reasoning about and perceiving something
non-human, and unknown or unfamiliar based on one's representation of the
familiar and well-known concept of being human (or one-self). The basic
operations underlying inductive inference are the acquisition of knowledge, the
activation or elicitation of knowledge, and the application of activated
knowledge at the time of the judgment \citep{epley_when_2008}. According to Epley
\textit{et al.}, the application includes attempts to correct, adjust, or
integrate less accessible information into a more automatically activated
default representation. This process can be seen as a process of correction, and
thus, it is often insufficient leaving final judgments biased in the direction
of the initially activated representation. As a person's knowledge base changes
/ evolves constantly with newly acquired things, or growing experiences with a
robot, etc., it is likely that the "need to anthropomorphize" a robot decreases
over time. First, because of the evolved knowledge about it, and second because
one has familiarized oneself with it. Consequently, the robot should become more
predictable and more familiar to the human user and inferences about it can be
made based on the acquired knowledge.  /fxfatal{confusing paragraph...needs
reformulation. Could also be turned into a section introduction instead of a
subsection}

So far, the HRI community has not much investigated how anthropomorphism in
human-robot interactions evolves over time (during the process of
\emph{adopting} a robot, for instance): we propose to go beyond the traditional
perception of anthropomorphism in robotics as a static feature that once
observed during a short-term interaction reflects a sustaining social effect.
Based on an extensive literature review previously
published~\citep{fink_anthropomorphism_2012} and field
experiments~\fixme{cites!}, we believe that anthropomorphic effects in HRI are
not always the same but evolve over time, along with growing interaction and
experience with the robot, similar to how relationships among people evolve
over time \fixme{(cite Hinde, 1988)}. In fact, \citet{kanda_interactive_2004}
also hypothesized that people's attitude toward technological artifacts and
their relationship with them would evolve over time, and they were among the
first who described \emph{novelty effects} during a long-term study with an
interactive robot in a school.

\hl{Julia: I need to work on this next paragraph but I would like to keep it here as a note}

\textbf{In-the-moment vs. reflective perspectives on agency}\\
Perception of agency can be treated as anthropomorphism.
\citep{takayama_perspectives_2012} proposes to differentiate between a \textit{in-the-moment} and \textit{reflective} perspective to better understand how human users perceive agency in robots.\footnote{``The other important distinction between the two perspectives is the \hl{cognitive processes} that they engage.''} According to Takayama, an \textit{in-the-moment} perspective refers to one's most immediate (sometimes visceral) sense in a situation. In contrast, a \textit{reflective} perspective refers to one's sense of a situation upon more distanced cogitation and consideration. For instance, we reflectively might not perceive agency in a social robot, but it can feel quite differently in the moment of interaction. A critical dimension of agency is the perspective from which something seems to have agency. \cite{takayama_perspectives_2012} argues that neglecting to separate reflective perspectives from in-the-moment perspectives of agency is one of the major sources of confusion when people talk and write about anthropomorphism. There seems to be a disconnection between what people consciously perceive and how they respond to stimuli that they may not consciously perceive.
That means, in the initial phase of interacting with a novel device \textbf{(initialization)} people might not consciously respond but rather mindlessly \citep{nass_machines_2000}. Only after some time of \textbf{familiarization}, they might respond in a more reflective manner. This can be illustrated by the fact that people tend to deny interacting with computational systems as if they were people and yet they respond to computers in many ways that are remarkably similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two different perspectives of in-the-moment \vs reflective to illustrate differences. She states that in-the-moment perceptions of agency are largely shaped by bottom-up perceptual processes, evoking very immediate responses (\eg to the cues emitted by the artifact's design, see Section \ref{sec:anthropomorphic-design}).
In contrast, according to Takayama, reflective perceptions are more often shaped by top-down processes because of the nature of reflective thought.

>>>>

\section{What does Anthropomorphism mean?}
\label{sec:anthropomorphism}

Anthropomorphism is an interesting phenomenon because it tells us something about 
the relationship between the human who anthropomorphizes, and the non-human agent 
which is anthropomorphized. Though commonly observed, it still appears controversial, at 
the same time, that humans tend to \textit{``see human in the non-
human''} \citep{epley_seeing_2007}. 
To shed light on this phenomenon, first of all, we would like to propose a clarification of terms.
In short, we propose to differentiate between \textbf{anthropomorphic form} to refer to an imitation of human-like form in the \textit{design} of artificial agents, and  \textbf{anthropomorphism} to refer to the social phenomenon of 
people's tendency to \textit{perceive} human-likeness in non human-like agents
\citep{fink_anthropomorphism_2012}. In the following, we might use the terms \texit{form} and \textit{design} interchangeably.

\hl{Julia: I will make a draft of a figure here}

As anthropomorphism arises in an interaction between a human and a non-human agent
\citep{persson_anthropomorphism_2000}, one can also
take two main perspectives on explaining anthropomorphism \citep{lee_human_2005}: an artifact-centered perspective and a human-centered perspective.\\ 
We now first present an artifact-centered perspective, and outline how 
robotics uses \textit{anthropomorphic design} to encourage 
people to interact in a social, human-like way with robots.
Then, we present a human-centered 
perspective and shed light on \textit{anthropomorphism as a social phenomenon}. We present 
theories from developmental psychology, cognitive science, and social psychology to explain people's tendency to 
perceive human-like characteristics in non-human agents.
It has to be kept in mind that both perspectives are inseparable; they apply at the 
same time, and we would like to present them here just as two different viewpoints 
on the same phenomenon.
In the end of this section, we synthesize 
the presented backgrounds into an extended understanding of anthropomorphism 
as a multi-factor phenomenon.

@@

On one hand, developers and designers have discovered that human-like robots may better 
match the affordances that this to humans adapted world brings along. 
Apart from this, it is suggested that an increased human-likeness of a robot (up to a 
certain threshold) will generally lead to 
more perceived \textit{affinity} from the human side \citep{mori_uncanny_1970} (see 
Figure \ref{fig:uncanny_valley}).
This further increases anthropomorphism, and as such
appears to be beneficial for human-robot relations, especially with socially 
interactive robots \citep{fong_survey_2003}. Consequently, the domain of (social) 
robotics tries to exploit this by using
anthropomorphic forms in robots.
Humans seem to readily accept and prefer the 
anthropomorphic form of a robot (up to a certain point), when compared to purely mechanical 
robots. A possible explanation for this is that perceiving a system as
human-like (or also pet-like) compared to machine-like changes the 
way how users try to understand and make sense of this system. The reason for this is 
that humans tend to project everyday 
expectations of human and animalistic life onto artificial agents that use 
anthropomorphic (or zoomorphic) forms \citep{schmitz_concepts_2011}. 
Both cognitive science 
and social psychology can provide explanations for this process of projection. The basic idea 
is that \textit{"the more similar in 
appearance, the more people are likely to use themselves as a source of induction 
and anthropomorphize these nonhuman agents"} \citep{epley_seeing_2007}. Studies have further 
shown that perceived similarity between the user and the robot influences the intensity of 
the participants' attributions towards the robot. Thus, the anthropomorphic design of a robot 
impacts the user's perception of the robot, and how far the user is likely to attribute and 
experience qualities such as intelligence, fun, or intentions in the robot. 
These experiences mark the transition between an immediate \textit{in-the-moment} and more thoughtful \textit{reflective} \citep{takayama_perspectives_2012} perception of, and interaction with the robot. As such, a more human-centered perspective needs to be applied, and we will come back to this in Section \ref{sec:human-centered}.

@@

Both the second factor \textit{effectance} and the third factor \textit{sociality} are \textbf{motivational determinants of anthropomorphism}. \textit{Effectance} is understood a person's motivation to interact effectively in one's environment. That means, a person is motivated to be able to understand, predict, and reduce uncertainty about her environment and the agents that inhabit it. According to \cite{epley_seeing_2007}, \textit{effectance motivation} can lead to anthropomorphism, since anthropomorphism serves as one way to reduce uncertainty and to increase comprehension of events in one's environment. Again, also according this factor it can be argued that children are more likely to anthropomorphize than adults. As children are in their early stages of life, they are likely to feel more uncertain within their environment, and consequently tend to anthropomorphize more.

@@

Summing it up, anthropomorphism is a commonly discussed and
studied trait of human-robot interaction; however its current understanding seems to be stuck in a twofold consideration. Mostly, only two key factors are recognized: the
characteristics of \textbf{1) the robot's design} (degree of human-likeness) and
the psychological determinants in \textbf{2) the human user} (see
\cite{epley_seeing_2007}). A lot of research has shown that both these factors can
promote or hinder anthropomorphism.
However, anthropomorphism is a complex process and we think that the current understanding of it does neither seem to fully take into account the wide range of phenomena that it encompasses, nor the multiplicity of factors which determine it.
This shortcoming and conflict in the understanding of anthropomorphism in HRI has already been denoted by others but we would like to take a first step towards moving forward in this dilemma, and encourage an open dialogue to solve this.

Through applying an understanding that anthropomorphism represents a specific type of \textit{experience} that arises in the
\textit{interaction} between a set of user expectations and external reality, 
%(see Figure~\ref{fig:anthropomorphism_and_interaction})
we propose that
anthropomorphism is a multi-layered phenomenon which arises in different levels \citep{persson_anthropomorphism_2000}, and which is likely to evolve with continued interaction.
%we suggest to understand anthropomorphism as a \emph{dynamic},
%\emph{context-dependent} process. To this end, we apply Persson \textit{et
%al.}'s six levels on anthropomorphism to HRI and introduce three
%\emph{interaction phases}. These interaction phases of anthropomorphism reflect
%also the results of a long-term study that we conducted in a real human
%environment and provide an account of the so-called \textit{novelty effect}.

To this end, and in respect of our research hypotheses, we make the following two suggestions:

@@

Second, we would like to take up the proposition already made by other researchers (\eg \cite{persson_anthropomorphism_2000,takayama_perspectives_2012,ruijten_introducing_2014}) to not viewing anthropomorphism as being either there or not there, but to discriminate different levels of
anthropomorphism (low, and high anthropomorphism, for instance).
This seems to make sense, since each of the four proposed factors has its own characteristics and involves specific
types of user expectations, for instance. Such a differentiation in the understanding of anthropomorphism could further allow to draw inferences
on the human-robot relationship based on the observed level of anthropomorphism.

@@

With having both before-mentioned aspects in mind (multi-factors that determine anthropomorphism and multi-levels of the observable phenomenon), it becomes clear that anthropomorphism cannot be something fixed, that once observed in a human-robot interaction always remains the same. We think, that anthropomorphism evolves and shows dynamics. This idea is what we would like to formalize in the next section.

@@

So far, the HRI community has not much investigated how anthropomorphism in
human-robot interactions evolves over time (during the process of
\emph{adopting} a robot, for instance). Though conflicts were found, the HRI community more or less accepted viewing anthropomorphism as a static feature that once
observed during a short-term interaction reflects a sustaining social effect. So far, it was also measured as such. But there are some hints that this is deficient.

For instance, \cite{kanda_interactive_2004} point out that relationships among people evolve over time and similarly, also people’s attitude toward technological artifacts
and their relationship with them might evolve over time. In their study, \cite{kanda_interactive_2004} were also among the first to describe empirically \emph{novelty effects} during a long-term study with an interactive robot in a school.

There are also hints from a more theoretical side, that make us think that anthropomorphism is likely to change over 
time and with growing experience that a person has with a robot.

%effectance motivation reduces uncertainty and consequently the tendency to anthropomorphize
For instance, let's have a look at the processes that underlie anthropomorphism.
In its essence anthropomorphism means reasoning about and perceiving something
non-human (and unknown or unfamiliar) based on one's representation of the
familiar and well-known concept of being human (or one-self). According to \citep{epley_when_2008} the basic
operations that underlie these inductive inferences are the acquisition of knowledge, the
activation or elicitation of knowledge, and the application of activated
knowledge at the time of the judgment. By applying knowledge to a given situation / interaction, it is implied that one attempts to correct, adjust, or maybe
integrate less accessible information into a more automatically activated
default representation. \cite{epley_when_2008} argues that this process can be seen as a process of correction, and
thus, it is often insufficient leaving final judgments biased in the direction
of the initially activated representation. Refinements to the initial representation mean something evolves. This can be due to the fact that a person's knowledge base changes
constantly with newly acquired things, or growing experiences with a
robot, \etc. . Consequently, it is likely that the ``need to anthropomorphize'' a robot decreases
over time (from the human-centered point of view), since the user's knowledge about the robot evolves, and also the user familiarizes herself with the robot. Consequently, the robot should become more
predictable and more familiar to the human user and inferences about it can be
made based on the acquired knowledge.

In other words, the argumentation is that humans have a motivation to interact effectively in their environment \textit{(effectance)} \citep{epley_when_2008}. This means that humans are motivated to be able to understand and predict the environment and agents around them. Through this process, uncertainty is reduced, and consequently, a first unfamiliar robot, can be explained with means that are different from anthropomorphism.

%expectations
Further,

%\textbf{In-the-moment vs. reflective perspectives on agency}\\
%Perception of agency can be treated as anthropomorphism.
%\citep{takayama_perspectives_2012} proposes to differentiate between a \textit{in-the-moment} and \textit{reflective} perspective to better understand how human users perceive agency in robots.\footnote{``The other important distinction between the two perspectives is the \hl{cognitive processes} that they engage.''} According to Takayama, an \textit{in-the-moment} perspective refers to one's most immediate (sometimes visceral) sense in a situation. In contrast, a \textit{reflective} perspective refers to one's sense of a situation upon more distanced cogitation and consideration. For instance, we reflectively might not perceive agency in a social robot, but it can feel quite differently in the moment of interaction. A critical dimension of agency is the perspective from which something seems to have agency. \cite{takayama_perspectives_2012} argues that neglecting to separate reflective perspectives from in-the-moment perspectives of agency is one of the major sources of confusion when people talk and write about anthropomorphism. There seems to be a disconnection between what people consciously perceive and how they respond to stimuli that they may not consciously perceive.
%That means, in the initial phase of interacting with a novel device \textbf{(initialization)} people might not consciously respond but rather mindlessly \citep{nass_machines_2000}. Only after some time of \textbf{familiarization}, they might respond in a more reflective manner. This can be illustrated by the fact that people tend to deny interacting with computational systems as if they were people and yet they respond to computers in many ways that are remarkably similar to how they respond to people \citep{reeves_media_1996}.
%\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two different perspectives of in-the-moment \vs reflective to illustrate differences. She states that in-the-moment perceptions of agency are largely shaped by bottom-up perceptual processes, evoking very immediate responses (\eg to the cues emitted by the artifact's design, see Section \ref{sec:anthropomorphic-design}).
%In contrast, according to Takayama, reflective perceptions are more often shaped by top-down processes because of the nature of reflective thought.
